Using Heavy-Tailed Self-Regularization Theory to Characterize Correlations in and Perform Diagnostics on State-of-the-art Overparameterized Models

### 2021-02-18 submission/TOPML_Workshop/paper.tex
edits added

In the HT-SR theory, one analyzes the eigenvalue spectrum, or empirical spectral density (ESD) 
of the associated layer correlation matrices~\citep{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.
and uses this to characterizes the amount and form of correlation present in the DNN's weight matrices.
For each layer weight matrix $\mathbf{W}$, if we compute the eigenvalue spectrum of 
$ \mathbf{X} = \frac{1}{N}\mathbf{W}^{T}\mathbf{W} $
then HT-SR theory characterizes the correlations in a weight matrix by fitting its ESD, $\rho(\lambda)$.
to a truncated Power Law (TPL) distribution,  $\rho(\lambda)\sim\lambda^{-\alpha} $.
%
The original work on HT-SR showed that among numerous publicly-available SOTA models, for nearly every $\mathbf{W}$,  the tail of the ESDs can be fit to a  TPL, and that Power Law (PL) exponents $\alpha$ nearly all lie within the range $\alpha\in(1.5,5)$~\citep{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.
%
%%As for the mechanism responsible for these properties, statistical physics offers several possibilities~\citep{SornetteBook,nishimori01}, e.g., self organized criticality~\citep{SOC87,SOCat25yrs} or multiplicative noise in the stochastic optimization algorithms used to train these models~\citep{HodMah20A_TR,SorCon97}.
%%Alternatively, related techniques have been used to analyze correlations and information propogation in actual spiking neurons~\citep{SYYRP11,YKYP14}.
%%Our meta-analysis does not require knowledge of mechanisms; and it is not even clear that one mechanism is responsible for every case.
%
Crucially, HT-SR theory predicts that smaller values of $\alpha$ should correspond to models with better correlation over multiple size scales and thus to better models,
and therefore higher generalization accuracies.
%%The notion of ``size scale'' is well-defined in physical systems, to which this style of analysis is usually applied, but it is less well-defined in CV and NLP applications.
%%Informally, it would correspond to pixel groups that are at a greater distance in some metric, or between sentence parts that are at a greater distance in text.
%%Relatedly, previous work observed that smaller exponents $\alpha$ correspond to more implicit self-regularization and better generalization, and that we expect a linear correlation between $\hat{\alpha}$ and model quality~\citep{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.

@@

We find that norm based metrics correlate well with reported test accuracies for well-trained models, but that they often cannot distinguish well-trained versus poorly-trained models.
We also find that PL based metrics can do much better---quantitatively better at discriminating among series of well-trained models with a given architecture; and qualitatively better at discriminating well-trained versus poorly-trained models.
These methods can be used to identify when a pretrained DNN has problems that cannot be detected simply by examining training/test accuracies.
%
%Here is a summary of our main results.
%First, norm-based metrics do a reasonably good job at predicting quality trends in well-trained CV/NLP models.
%Second, norm-based metrics may give spurious results when applied to poorly-trained models (e.g., models trained without enough data, etc.).
%For example, they may exhibit what we call Scale Collapse for these models. 
%Third, PL-based metrics can do much better at predicting quality trends in pretrained CV/NLP models.  
%In particular, 
%a weighted PL exponent (weighted by the log of the spectral norm of the corresponding layer) is 
%quantitatively better at discriminating among a series of well-trained versus very-well-trained models
%within a given architecture series; and
%the (unweighted) average PL exponent is 
%qualitatively better at discriminating well-trained versus poorly-trained models.
%Fourth, PL-based metrics can also be used to characterize fine-scale model properties, including what we call layer-wise Correlation Flow, in well-trained and poorly-trained models; and they can be used to evaluate model enhancements (e.g., distillation, fine-tuning, etc.).
See Table~\ref{table:results} for an overview summary of our results.

>>>>

In the HT-SR theory, one analyzes the eigenvalue spectrum, or empirical spectral density (ESD) 
of the associated layer correlation matrices~\citep{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.
and uses this to characterizes the amount and form of correlation present in the DNN's weight matrices.
For each layer weight matrix $\mathbf{W}$, if we compute the eigenvalue spectrum of 
$ \mathbf{X} = \frac{1}{N}\mathbf{W}^{T}\mathbf{W} $
then HT-SR theory characterizes the correlations in a weight matrix by fitting its ESD, $\rho(\lambda)$.
to a truncated Power Law (TPL) distribution,  $\rho(\lambda)\sim\lambda^{-\alpha} $.
%
The original work on HT-SR showed that among numerous publicly-available SOTA models, for nearly every $\mathbf{W}$,  the tail of the ESDs can be fit to a  TPL, and that
the Power Law (PL) exponents $\alpha$ nearly all lie within the range $\alpha\in(1.5,5)$~\citep{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.
%
%%As for the mechanism responsible for these properties, statistical physics offers several possibilities~\citep{SornetteBook,nishimori01}, e.g., self organized criticality~\citep{SOC87,SOCat25yrs} or multiplicative noise in the stochastic optimization algorithms used to train these models~\citep{HodMah20A_TR,SorCon97}.
%%Alternatively, related techniques have been used to analyze correlations and information propogation in actual spiking neurons~\citep{SYYRP11,YKYP14}.
%%Our meta-analysis does not require knowledge of mechanisms; and it is not even clear that one mechanism is responsible for every case.
%
Crucially, HT-SR theory predicts that smaller values of $\alpha$ should correspond to models with better correlation over multiple size scales and thus to better models,
and therefore higher generalization accuracies.
%%The notion of ``size scale'' is well-defined in physical systems, to which this style of analysis is usually applied, but it is less well-defined in CV and NLP applications.
%%Informally, it would correspond to pixel groups that are at a greater distance in some metric, or between sentence parts that are at a greater distance in text.
%%Relatedly, previous work observed that smaller exponents $\alpha$ correspond to more implicit self-regularization and better generalization, and that we expect a linear correlation between $\hat{\alpha}$ and model quality~\citep{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.

@@

We find that norm based metrics (may) correlate well with reported test accuracies for well-trained models, 
but that they often cannot distinguish well-trained versus poorly-trained models.
We also find that PL based metrics can do much better---quantitatively better at discriminating among series of well-trained models with a given architecture; and qualitatively better at discriminating well-trained versus poorly-trained models.
These methods can be used to identify when a pretrained DNN has problems that cannot be detected simply by examining training/test accuracies.
%
%Here is a summary of our main results.
%First, norm-based metrics do a reasonably good job at predicting quality trends in well-trained CV/NLP models.
%Second, norm-based metrics may give spurious results when applied to poorly-trained models (e.g., models trained without enough data, etc.).
%For example, they may exhibit what we call Scale Collapse for these models. 
%Third, PL-based metrics can do much better at predicting quality trends in pretrained CV/NLP models.  
%In particular, 
%a weighted PL exponent (weighted by the log of the spectral norm of the corresponding layer) is 
%quantitatively better at discriminating among a series of well-trained versus very-well-trained models
%within a given architecture series; and
%the (unweighted) average PL exponent is 
%qualitatively better at discriminating well-trained versus poorly-trained models.
%Fourth, PL-based metrics can also be used to characterize fine-scale model properties, including what we call layer-wise Correlation Flow, in well-trained and poorly-trained models; and they can be used to evaluate model enhancements (e.g., distillation, fine-tuning, etc.).
See Table~\ref{table:results} for an overview summary of our results.
